---
title: "Lab 5 assignment - Point pattern analysis"
output:
  html_document:
    df_print: paged
---

```{r, warning=F, error=F, message=F}
require(spatstat)
require(tidyverse)
require(terra)
```


# Challenge 1 (3 points)

Create a simulation window on a landscape stretching from 0 to 100 units in both the x and y directions. Now simulate the distribution of an organism on this landscape using a Poisson process with a lambda value of 0.01. Plot this organism's distribution. How many organisms are there on your landscape? What is the expected number of organisms that you should find on this landscape based on your simulation parameters? Why?

```{r}
window <- owin(xrange=c(0,100), yrange=c(0,100))

set.seed(121)
pop <- rpoispp(lambda=0.01, win=window)
plot(pop)
```

I have 97 organisms. Lambda is points per unit area. There are 10,000 units * intensity of 0.01 = 100 expected organisms.

# Challenge 2 (3 points)

Verify that there no linear or quadratic pattern in the distribution of your organisms using fitted Poisson process models. Show the outcome of this analysis and explain your conclusion.

```{r}
int <- ppm(pop, ~1)
linear <- ppm(pop, ~x+y)
quad <- ppm(pop, ~polynom(x,y,2))

data.frame(model <- c('ident', 'linear', 'quad'),
           AIC <- c(AIC(int), AIC(linear), AIC(quad)))
```

Linear and quadratic models have higher AIC than a model with just 1, so those patterns are not present.

# Challenge 3 (14 points)

I have simulated the distributions of 2 critters and provided their x-y coordinates. One of them is a non-territorial predator. The second is a critter that is preyed upon by the predator and exhibits a fear response to the predator's presence. Use the tools you learned in the lab to explore the distributions of these two critters. Identify which one is the predator and which is the prey, and provide code and a thorough justification supporting your conclusion.


```{r}
#Use this code chunk to get you started.

predPreyWin = owin(xrange = c(0, 100), yrange = c(0, 100))

critter1 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species1.csv') %>% 
  mutate(org = 'critter1')

critter2 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species2.csv') %>% 
  mutate(org = 'critter2')

ggplot() +
  geom_point(data=critter1, aes(x=x, y=y)) +
  geom_point(data=critter2, aes(x=x, y=y), color='red') +
  theme_bw() +
  theme(aspect.ratio=1)

```

Integrate your answer with additional code chunks here.

```{r}

points <- rbind(critter1, critter2)

pppCritter <- ppp(x=points$x, y=points$y, window=predPreyWin, marks=factor(points$org))

Lcross <- Lcross(pppCritter, "critter1", "critter2")
plot(Lcross)

Lcross2 <- Lcross(pppCritter, "critter1", "critter1")
plot(Lcross2)

Lcross3 <- Lcross(pppCritter, "critter2", "critter2")
plot(Lcross3)

```


In `Lcross`, `Critter1` and `Critter2` are consistently further from each other than expected. In `Lcross2`, `Critter1` is slightly clumped. In `Lcross3`, `Critter2` is slightly clumped between 10-20 m but slightly spaced out around 5 m.

```{r}

pppC1 <- ppp(x=critter1$x, y=critter1$y, window=predPreyWin)
pppC2 <- ppp(x=critter2$x, y=critter2$y, window=predPreyWin)

q1 <- quadratcount(pppC1, nx=4, ny=4)
plot(pppC1)
plot(q1, add=T)

q2 <- quadratcount(pppC2, nx=4, ny=4)
plot(pppC2)
plot(q2, add=T)


quadrat.test(pppC1, nx=4, ny=4, method='Chisq')
quadrat.test(pppC2, nx=4, ny=4, method='Chisq')

```


Chi-squared tests of CSR show that Critter 1 is significantly clumped at this scale while Critter 2 is not (alpha = 0.05).

```{r}

lCsr1 <- envelope(pppC1, Lest, nsim=99, rank=1, correction='translate', global=F)
plot(lCsr1, .-r~r, shade=c('hi', 'lo'), legend=F)

lCsr2 <- envelope(pppC2, Lest, nsim=99, rank=1, correction='translate', global=F)
plot(lCsr2, .-r~r, shade=c('hi', 'lo'), legend=F)

lCritter <- envelope(pppCritter, Lest, nsim=99, rank=1, i='org', global=F)
plot(lCritter, .-r~r, legend=F)

pEnv1 <- envelope(pppC1, pcf, nsim=99, rank=1, correction='translate', global=F)
plot(pEnv1, shade=c('hi', 'lo'), legend=F)

pEnv2 <- envelope(pppC2, pcf, nsim=99, rank=1, correction='translate', global=F)
plot(pEnv2, shade=c('hi', 'lo'), legend=F)

```


I think these analyses suggest that Critter 1 is the prey and Critter 2 is the predator. Critter 1 are congregating in predator-free areas at small spatial scales under ~11 m. `lCritter` shows Critter 1's distribution relative to Critter 2's, which means that under 3 m they are close, but above 3 m they are more separated than expected; suggesting this is the distance that they flee the predator. Critter 2 is slightly more distributed at smaller scales under 3 m, which would happen if they are pursuing prey in different directions.